---
title: "Exploratory analysis on album dataset"
output:
  html_document: default
  pdf_document: default
date: "2024-01-23"
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

### This project explors the "album.csv" dataset. It works with music genres, artists and music albums. 

## Data preprocessing 

First, I load the necessary libraries (install packages, id needed), read the data set and print its content!
```{r csv_file_path}
# Install and load the tidyverse package if not already installed
 # install.packages("tidyverse")
 # install.packages("ggplot2")
 # install.packages("spData")
library(tidyverse)
library(stringr)
library(spData)
library(sf)
library(tmap)

# Set the path to your CSV file
csv_file_path <- "album.csv"

# Read the CSV file
data <- read.csv(csv_file_path)

# Print the contents of the data
#print(data)

```



Then, the dataset is processed, such that each row corresponds to one genre for each artist. Each artist corresponds to as many rows as the genres they participate to.

```{r data_processed}
data_processed <- data %>%
  # Split the 'genre' column into a list of genres
  separate_rows(genre, sep = ",") %>%
  # Remove leading and trailing whitespaces
  mutate(across(c(genre, artist, album), str_trim))

#print(data_processed)



data_processed_new <- data_processed %>%
  # Extract year from the 'album' column
  mutate(year = str_extract(album, "\\((\\d{4})\\)"),
         album = str_replace(album, "\\(\\d{4}\\)", "")) %>%
  # Remove leading and trailing whitespaces
  mutate(across(c(genre, artist, album), str_trim))

# Print the new processed data with the new "year" column
#print(data_processed_new)
```


The final step of the pre-processing involves the creation of "artists" , as required for question 4.

```{r artists}
#Create "artists" for Question 4 

# Create 'artists' data frame
artists <- data_processed_new %>%
  group_by(artist) %>%
  summarize(
    primary_genre = first(genre),
    secondary_genre = ifelse(n() > 1, nth(genre, 2), NA_character_),
    .groups = 'drop'
  )


# Replace NAs in secondary_genre with corresponding values from primary_genre
artists <- artists %>%
  mutate(secondary_genre = ifelse(is.na(secondary_genre), primary_genre, secondary_genre))


# Print the 'artists' data frame
#print(artists)


# Print the "Wim Mertens" row as a test
artist_data_filtered <- artists %>%
  filter(artist == "Mertens, Wim")

# Print the filtered data
print(artist_data_filtered)

```


## Artist Percentage per Genre

```{r genre_percentage}
# Calculate the percentage of artists for each genre
genre_percentage <- data_processed_new %>%
  group_by(genre, artist) %>%
  summarize(Base_Artist_Genre = first(genre)) %>%
  group_by(Base_Artist_Genre) %>%
  summarize(Percentage = n() / n_distinct(data_processed_new$artist) * 100)

# Print the results
#print(genre_percentage)


# Create a bar chart
ggplot(genre_percentage, aes(x = Base_Artist_Genre, y = Percentage, fill = Base_Artist_Genre)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of Artists in Each Genre",
       x = "Genre",
       y = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Album Distribution per Year

```{r data_processed_new}
# Remove non-numeric characters from the "year" column and convert to numeric
data_processed_new$year <- as.numeric(gsub("\\D", "", data_processed_new$year))

# Plot the histogram
ggplot(data_processed_new, aes(x = year)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Albums per Year",
       x = "Year",
       y = "Number of Albums") +
  theme_minimal()



# Plot the histogram with normalized y-axis
ggplot(data_processed_new, aes(x = year)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7, aes(y = ..density..)) +
  labs(title = "Normalised Distribution of Albums per Year",
       x = "Year",
       y = "Proportion of Albums") +
  theme_minimal()


```


## Album Distribution for top 5 genres, per Year

```{r selected_data}
# Select the relevant columns and filter out NAs in the year column
selected_data <- data_processed_new %>%
  select(genre, year) %>%
  filter(!is.na(year))

# Count the occurrences of each genre in each year
genre_counts <- selected_data %>%
  group_by(year, genre) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

# Identify the top 5 genres
top_genres <- genre_counts %>%
  group_by(genre) %>%
  summarize(total_count = sum(count)) %>%
  top_n(5, total_count) %>%
  pull(genre)

# Filter the data for the top 5 genres
top_genre_data <- genre_counts %>%
  filter(genre %in% top_genres)

# Plot the distribution of the top 5 genres over time
ggplot(top_genre_data, aes(x = year, y = count, color = genre)) +
  geom_line() +
  labs(title = "Distribution of Top 5 Genres Over Time",
       x = "Year",
       y = "Number of Albums") +
  theme_minimal()



```


## Exploring Relationships among genres - Two Way Table

This part creates a mosaic plot, using a two way table ("artists") in order to visualise the relationships between the primary and secondary genres!

```{r two_way, fig.width=10, fig.height=8}
two_way <- table(artists$primary_genre, artists$secondary_genre)
summary(two_way)
mosaicplot(two_way ,main="Relationship Between Primary and Secondary Genres",las=2)


```





First, I create the new column "country_in_title" in "data_processed_new "with the country name that appears on each album title (or NA - if no country is included on the title)

```{r world}
# Read world data set, print and plot it!
world <- st_read(system.file("shapes/world.gpkg", package="spData"))
#print(world)
plot(world)


# Extract unique country names from the world data set
country_names <- world$name_long

# Create a new column in data_processed_new for the first country name found in each album title
data_processed_map <- data %>%
  mutate(country_in_title = sapply(album, function(title) {
    found_country <- country_names[str_detect(title, country_names)]
    if (length(found_country) > 0) found_country[1] else NA
  }))


```


Then, I calculate how many times each country appears on album titles and create the choropleth map with this information!

First map: Using binning!
```{r map}
#Aggregate the data to count the frequency of each country
country_frequency <- data_processed_map %>%
  group_by(country_in_title) %>%
  summarize(frequency = n()) %>%
  filter(!is.na(country_in_title))

# Load the world shapefile
world_shapefile <- st_read(system.file("shapes/world.gpkg", package="spData"))

# Merge the data with the world shapefile
map_data <- left_join(world_shapefile, country_frequency, by = c("name_long" = "country_in_title"))

# Plot the choropleth map
tm_shape(map_data) +
  tm_borders(lwd = 0.5) +
  tm_fill("frequency", style = "jenks", title = "Frequency", palette = "Blues") +
  tm_layout(title = "Choropleth Map (with binning) of Country Frequencies in Song Titles") +
  tm_legend(legend.title.size = 1.5, position = c("right", "bottom"))



```

Second map: Using a color scale!
```{r scale}
#Plot the choropleth map with a color scale
tm_shape(map_data) +
  tm_borders(lwd = 0.5) +
  tm_fill("frequency", style = "cont", title = "Frequency", palette = "Blues") +
  tm_layout(title = "Choropleth Map(with scale) of Country Frequencies in Song Titles") +
  tm_legend(legend.title.size = 1.5, position = c("right", "bottom"))




``` 

```{r }

# PLOT USING GGPLOT
#library(tidyverse)
#library(sf)
#library(ggplot2)

# Aggregate the data to count the frequency of each country
# country_frequency <- data_processed_new %>%
#   group_by(country_in_title) %>%
#   summarize(frequency = n()) %>%
#   filter(!is.na(country_in_title))
# 
# # Load the world shapefile
# world_shapefile <- st_read(system.file("shapes/world.gpkg", package="spData"))
# 
# # Merge the data with the world shapefile
# map_data <- left_join(world_shapefile, country_frequency, by = c("name_long" = "country_in_title"))
# 
# # Plot the choropleth map using ggplot2
# ggplot(map_data, aes(fill = frequency)) +
#   geom_sf(color = "white", size = 0.2) +
#   scale_fill_viridis_c(name = "Frequency", option = "plasma", na.value = "white") +
#   theme_minimal() +
#   labs(title = "Choropleth Map of Country Frequencies in Song Titles")


```


## Shinny App! 


### a) Input: Dropdown list with artists, Output: Chosen artist's albums!


```{r shiny-app-a}
library(shiny)

# Define UI for application
ui <- fluidPage(
  titlePanel("Album Explorer"),
  
  # Dropdown for selecting artist
  selectInput("artist", "Select Artist:", choices = unique(data_processed_new$artist)),
  
  # Table to display albums
  tableOutput("album_table")
)

# Define server logic
server <- function(input, output) {
  # Filter data based on selected artist
  selected_data <- reactive({
    filter(data, artist == input$artist)
  })
  
  # Render the table
  output$album_table <- renderTable({
    selected_data()
  })
}

# Run the application
shinyApp(ui = ui, server = server)


```


### b) Input: Dropdown list with genres, Output: Chosen genre's albums (alphabetically)!

```{r shiny-app-b}

# Define the Shiny app
shinyApp(
  # Define UI components
  ui = fluidPage(
    titlePanel("Albums by Genre"),
    sidebarLayout(
      sidebarPanel(
        # Dropdown for selecting genre
        selectInput("genre", "Select Genre", choices = unique(data_processed_new$genre))
      ),
      mainPanel(
        # Table to display albums
        tableOutput("albumTable")
      )
    )
  ),
  
  # Define server logic
  server = function(input, output) {
    # Filter data based on selected genre
    filtered_data <- reactive({
      data_processed_new %>%
        filter(str_detect(genre, input$genre)) %>%
        arrange(album)  # Arrange alphabetically by album
    })

    # Output the table of albums
    output$albumTable <- renderTable({
      filtered_data()
    })
  }
)
```

### c) Input: Dropdown list with genres, Output: Chosen genre's artists (alphabetically)!

```{r shiny-app-c}
# Define the Shiny app
shinyApp(
  # Define UI components
  ui = fluidPage(
    titlePanel("Artists by Genre"),
    sidebarLayout(
      sidebarPanel(
        # Dropdown for selecting genre
        selectInput("genre_artist", "Select Genre", choices = unique(data_processed_new$genre))
      ),
      mainPanel(
        # Table to display artists
        tableOutput("artistTable")
      )
    )
  ),
  
  # Define server logic
  server = function(input, output) {
    # Filter data based on selected genre for artists
    filtered_data_artist <- reactive({
      artists %>%
        filter(str_detect(primary_genre, input$genre_artist) | str_detect(secondary_genre, input$genre_artist)) %>%
        arrange(artist)  # Arrange alphabetically by artist
    })

    # Output the table of artists
    output$artistTable <- renderTable({
      filtered_data_artist()
    })
  }
)

```

### d) Input: Dropdown list with genres, Output: Plot of the Distribution of this genres' albums accross time!

```{r shiny-app-d}
# Define the Shiny app
shinyApp(
  # Define UI components
  ui = fluidPage(
    titlePanel("Album Distribution Over Time"),
    sidebarLayout(
      sidebarPanel(
        # Dropdown for selecting genre
        selectInput("genre_time", "Select Genre", choices = unique(data_processed_new$genre))
      ),
      mainPanel(
        # Plot to display distribution over time
        plotOutput("albumDistributionPlot")
      )
    )
  ),
  
  # Define server logic
  server = function(input, output) {
    # Filter data based on selected genre for album distribution over time
    filtered_data_time <- reactive({
      selected_data <- data_processed_new %>%
        filter(str_detect(genre, input$genre_time)) %>%
        filter(!is.na(year))  # Remove entries without a year
    })

    # Output the plot of album distribution over time
    output$albumDistributionPlot <- renderPlot({
      ggplot(filtered_data_time(), aes(x = year)) +
        geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
        labs(title = paste("Distribution of", input$genre_time, "Albums Over Time"),
             x = "Year",
             y = "Number of Albums") +
        theme_minimal()
    })
  }
)


```


